---
title: "Survival Analysis Modeling in R"
output: html_notebook
author: "Clay Ford"
editor_options: 
  chunk_output_type: inline
---

This is an R Markdown Notebook. When you execute code within the notebook, the results appear beneath the code. 

Try executing this chunk by clicking the *Run* button within the chunk or by placing your cursor inside it and pressing *Ctrl+Shift+Enter* (Win/Linux) or *Cmd+Shift+Return* (Mac). 

```{r}
plot(cars)
```

To hide the output, click the Expand/Collapse output button. To clear results (or an error), click the "x" or simply run corrected code. 

You can also press *Ctrl+Enter* (Win/Linux) or *Cmd+Return* (Mac) to run one line of code at a time (instead of the entire chunk).

Add a new R code chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I* (Win/Linux) or *Cmd+Option+I* (Mac).  

## CODE ALONG 0

Enter a new code chunk and run the code `rnorm(10)` (sample 10 random values from a standard normal distribution.)


## Introduction

Survival analysis is a collection of procedures for analyzing the _time until an event occurs_. The methods were mostly developed in the medical field and concerned time until death, hence the name "survival analysis". Also known as _duration analysis_, _reliability analysis_, or _time-to-event analysis_. In this workshop we'll use the term "survival" generically.

What makes survival analysis challenging is that rarely do all subjects experience the event. When a subject does not experience the event, they are _censored_. 

A study that is analyzed using survival analysis typically has an _observation period_ during which subjects are enrolled and observed. When a subject is censored, that means they didn't experience the event _during_ the observation period. It does not mean they will never experience the event.

In this workshop we want to get you up and running with _modeling survival data_. This is where we examine the relationship between survival time and predictors of interest. This is similar to traditional regression modeling, but is made complicated due to censoring.

## Load packages and data

We're going to use the following packages in this workshop. 

```{r}
library(survival)
library(survminer)
```

Let's load some data. Two groups of leukemia patients. Patients in remission were randomly assigned to maintenance therapy with either 6-MP or placebo. 21 on treatment, 21 on placebo. (Freireich et al, Blood (1963) 21 (6): 699–716.) The event of interest is "coming out of remission." Going forward when we talk about "survival" or "failure", we mean the event "coming out of remission".

```{r}
d <- readRDS(url("https://github.com/clayford/survival_analysis_modeling/raw/main/data/leuk.rds"))
```

Variables:  
- survt (time in weeks; time until failure or censoring)
- status (status; 0 = censored, 1 = failure)
- sex (0 = female, 1 = male)
- logwbc (log transformed white blood cell count)
- rx (treatment or placebo)
- lwbc3 (log WBC categorized in 3 ordered categories)

```{r}
head(d, n = 8)
```

It's important to note that binary "status" doesn't mean "success/failure" like it does in data analyzed with binary logistic regression. It indicates whether we observed the event of interest _during the observation period_.

Some questions of interest: 

- Does the 6-MP treatment increase time until failure? 
- Do we need to control for log WBC? 
- Is the treatment effect the same for males and females?

## Descriptive methods for survival data

Look at censoring by treatment group. All subjects come out of remission in the placebo group (21). 12 subjects were censored in the trt group. That is, they did not experience failure during the study.

```{r}
xtabs(~ rx + status, data = d)
```


We could also look at average survival times (ie, time observed until the event or censoring). Subjects on "trt" appeared to survive longer. But because there are several censored subjects in the trt group, the true survival time is even larger.

```{r}
aggregate(survt ~ rx, data = d, mean)
```

An alternative statistics is the _hazard rate_, defined by total number of failures divided sum of survival times.

```{r}
# failures (ie, events) per rx group
f <- xtabs( ~ rx, data = d, subset = status == 1)
# total survival time by rx group
total <- tapply(d$survt, d$rx, sum)
# hazard rate by rx group
hr <- f/total
hr
```

Higher hazard rate means lower probability of survival. A ratio of hazards is called the _hazard ratio_.

```{r}
hr[2]/hr[1]
```

It appears hazard of remission is about 4 times higher in the placebo group. However these are just overall comparisons. We prefer to describe the data _over time_.

The _Kaplan-Meier (KM) curve_ is the workhorse function for visualizing and describing survival data over time. This method was published by Edward Kaplan and Paul Meier in 1958. We can create a KM curve using the `survfit()` function from the {survival} package. To do this, we need to define survival time data using the `Surv()` function. The first argument is survival time, the second argument is the censoring indicator. Values with "+" appended are censored values.

```{r}
Surv(d$survt, d$status)
```

The `survfit()` function has `formula` and `data` arguments, so we can directly use the variable names in the `Surv()` function. The `~ 1` says we want to fit a single survival curve for the entire data set.

```{r}
fit <- survfit(Surv(survt, status) ~ 1, data = d)
fit
```

n: number of subjects  
events: number of subjects who experienced the event  
median: median survival time  
0.95LCL, 0.95UCL: lower/upper 95% CI on median survival time

Using `plot()` we can create a KM survival curve with 95% confidence bands (dashed lines). The y-axis is probability of "survival". Each "step" down represents 1 or more subjects experiencing the event. There appears to be about a 0.8 probability of not experiencing a remission to week 5.

```{r}
plot(fit)
```

Calling summary on the KM objects returns survival probabilities (probability of failure, or coming out of remission) for each observed point in time.

```{r}
summary(fit)
```

Can specify specific times in `summary()` using the `times` argument. Below we look at weeks 5, 10, 15, and 20

```{r}
summary(fit, times = seq(5,20,5))
```

We're probably more interested in time-until-failure by rx group. For this we specify formula `Surv(survt, status) ~ rx`. Now we see different estimated median survival times by rx group.

```{r}
fit2 <- survfit(Surv(survt, status) ~ rx, data = d)
fit2
```

Estimated median time-to-failure for subjects in "trt" group is 23 weeks versus 8 weeks for the "placebo" group. 95% confidence intervals on the median estimates are provided. Notice the upper limit is NA for the "trt" group. This is because less than half the subjects in this group came out of remission.

This fit generates two KM curves. There is a `plot()` method but it looks pretty basic without some extra work. At a minimum we need to use the `lty` (line type) argument to distinguish between groups

```{r}
plot(fit2, lty = 1:2)
```

The `ggsurvplot()` function from the {survminer} package looks nicer and includes "+" symbols to indicate censored subjects and a legend to identify groups.

```{r}
ggsurvplot(fit2)
```

Setting `risk.table = TRUE` adds a table below the plot showing numbers of subjects at risk at various times in the observation period. The `tables.height = 0.3` argument specifies the height of the table. I arrived at this value using trial and error.

```{r}
ggsurvplot(fit2, risk.table = TRUE, tables.height = 0.3)
```

Finally we might be interested in testing the null hypothesis that both survival curves take the same trajectory. To answer this we can use the _log-rank test_, which is implemented in the `survdiff()` function. A small p-value provides evidence against the null hypothesis. Below we have good evidence the KM curves differ by rx group.

```{r}
survdiff(Surv(survt, status) ~ rx, data = d)
```

We usually want to analyze time-to-event _adjusting for additional variables_, such as age, lab values, patient history, etc. For this we turn to regression modeling.

## CODE ALONG 1

For our code alongs we'll work with data from the Worcester Heart Attack Study (WHAS). The main goal of this study was to describe factors associated with survival time following hospital admission for acute myocardial infarction, aka heart attack. The time variable is "time" (in years) and the censoring variable is "fstat".

```{r}
whas <- readRDS(url("https://github.com/clayford/survival_analysis_modeling/raw/main/data/WHAS.rds"))
```

(1) Create a KM curve for time versus "gender" (0 = male, 1 = female)


(2) What are the estimated survival probabilities for years 1 - 5?


(3) Run a log-rank test on the KM curves for gender



## Regression models for survival data

Regression models allow us to study the effect of multiple variables on the survival experience. For modeling survival time it is convenient to model the _hazard of the event_ instead of the raw time-until-event or probability of event. This means we're estimating a _hazard function_.

It turns out that once we estimate a hazard function, we can also estimate a survival function, which means we can use our model to produce _adjusted survival curves_ similar to the KM curves we created to describe the data. 

The standard regression model for survival data is the _Cox Proportional Hazards Model_, named for David Cox who published the method in 1972. The Cox Proportional Hazards Model makes _no assumption_ about the distribution of the time-to-event. It is usually a safe and robust model. Even if there is a better-fitting "correct" parametric model (i.e., a model that makes assumptions about the distribution of survival time), the Cox model will often closely approximate the results.

The {survival} package provides the `coxph()` function for fitting Cox PH models. We use the function as we would `lm()` or `glm()`, but with the left-hand side requiring the use of the `Surv()` function.

Below we model time-to-remission as a function of rx and save as `m1` and then use `summary()` to request the model summary.

```{r}
m1 <- coxph(Surv(survt, status) ~ rx, data = d)
summary(m1)
```

The coefficient we're usually most interested in is the one labeled "exp(coef)". This is the _hazard ratio_. The interpretation above is that the hazard of failure for those on "placebo" is about 4.8 times higher than the hazard of failure for those on "treatment".

- "se(coef)" is the _standard error_ of the coefficient.
- "z" is the _test statistic_ of the coefficient. z = coef/se(coef)
- "Pr(>|z|)" is the _p-value_, or probability of getting a z value bigger than the z observed (in absolute value) if the coefficient is truly 0. A small p-value provides evidence against the null hypothesis of the coefficient being 0.

The next section shows a 95% confidence interval on the hazard ratio. The hazard of failure for those on placebo is plausibly anywhere from 2 to 10 times higher than the hazard of failure for the trt group.

The "exp(-coef)" is the hazard ratio in the opposite direction. The hazard of failure on "treatment" is about 80% lower than the hazard of failure on "placebo".

The statistic labeled "Concordance" reports the fraction of all pairs of subjects where the model correctly predicts the individual with the earlier event. Values in the 0.50 - 0.55 range suggest a flip of a coin would do as well as our model at correctly ordering pairs of subjects. The value of 0.69 is promising but not excellent. 

The three tests at the end each test the null that all model coefficients are simultaneously equal to 0. These are usually all in agreement. When these disagree, go with the Likelihood Ratio Test (Hosmer et al, p. 79)

We can use the `ggadjustedcurves()` function from the {survminer} package to plot adjusted survival curves from our model.

```{r}
ggadjustedcurves(m1, variable = "rx", data = d)
```

Compare to the descriptive KM plot (generated below). Notice the adjusted survival curves from the Cox PH model fit do not indicate censored observations. These are model-predicted expected survival curves calculated based on Cox PH model. The KM plot shows observed data. Also notice the model-generated plot stops around 20 weeks and does not extrapolate since no one in the placebo group lasted beyond then.

```{r}
# KM plot
ggsurvplot(fit2)
```

## CODE ALONG 2

(1) Fit a Cox PH model for time on gender using the WHAS data. Is this a good model for understanding survival time after admission to hospital for a heart attack?


(2) How do we interpret the coefficient for gender?


(3) Plot the adjusted survival curves for the Cox PH model



## The proportional hazards assumption

The Cox PH model assumes that the _hazard ratio is constant over time_. The model above implies the hazard of failure on placebo is about 4.8 times higher than the hazard of failure on treatment, no matter how long we observe the subjects. Notice the summary output above makes no reference to time.

Fortunately the {survival} package makes it pretty easy to assess these assumptions using the `cox.zph()` function. Use it on your model object. The null hypothesis is the hazard ratios are independent of time. A small p-value is evidence against the null hypothesis.

```{r}
cox.zph(m1)
```

Each variable is tested. The GLOBAL test is that all variables simultaneously satisfy the proportional hazards assumption. In this case they're the same since we only have one predictor. Set `global=FALSE` to suppress the global test.

A visual check is available using the `plot()` method. A fairly straight line indicates the proportional hazards assumption is likely satisfied.

```{r}
plot(cox.zph(m1))
```

If the proportional hazards assumption appears to be violated, we may need to consider fitting a _stratified Cox PH model_ (among other options). We discuss this later in the workshop.


## CODE ALONG 3

(1) Check the proportional hazards assumption of the model we fit in CODE ALONG 2 using a test.


(2) Visually check the proportional hazards assumption of the model we fit in CODE ALONG 2.


## Cox PH model with multiple predictors

In our previous model we only looked at the effect of "rx" on time to failure. However we may also want to adjust for white blood cell count ("logwbc"). This is a known prognostic indicator of coming out of remission for leukemia patients. Indeed we can see an association between "survt" (time-to-failure) and "logwbc", for those subjects who experienced the event (status = 1).

```{r}
plot(survt ~ logwbc, data = d, subset = status == 1)
```

Below we add the variable "logwbc" (log-transformed white blood cell count) to the model. This allows us to estimate placebo and treatment effects on patients with similar white blood cell counts. 

```{r}
m2 <- coxph(Surv(survt, status) ~ rx + logwbc, data = d)
summary(m2)
```

Adjusted for white blood cell count, it appears the hazard of failure for subjects on placebo is about 4 times higher than the hazard of failure for subjects on treatment. Likewise, adjusted for rx, the hazard of failure increases by about a factor of 5 for each one unit increase in logwbc. (In this case, that's probably not an interpretation we're interested in. We simply want to adjust for subjects' white blood cell count.)

Adjusting for logwbc reduces the effect of rx about 12%. This is a sign that logwbc may be _confounded_ with rx. 

```{r}
coef(m2)[1]/coef(m1)[1]
```

The confidence interval on the rx hazard ratio is also tighter (more precise) in the model with logwbc.

Model 1 without logwbc
CI width (upper bound - lower bound):    
10.81 - 2.147 = 8.663

Model 2 with logwbc
CI width (upper bound - lower bound):  
9.195 - 1.739 = 7.456

We can formally compare models using the `anova()` function. The null hypothesis of this test is that the models are equally adequate. The test below suggests we should keep logwbc in our model.

```{r}
anova(m1, m2)
```

We should assess the proportional hazards assumption again. Recall this assumption applies to all predictors in the model.

```{r}
cox.zph(m2)
```


```{r}
plot(cox.zph(m2))
```

Once again we can generate adjusted survival curves using our model. 

```{r}
ggadjustedcurves(m2, variable = "rx", data = d)
```

The technical details of how adjusted survival curves are calculated is beyond the scope of our presentation. By default, the `ggadjustedcurves()` function creates _conditional_ survival curves. Basically, all our data is used twice to predict expected survival: once as if everyone were in the "placebo" group, and again as if everyone were in the "treatment" group. See the help page for `ggadjustedcurves()` and the vignette that accompanies the {survival} package, "Adjusted Survival Curves". 

To see the values that are plotted, we can call the `surv_adjustedcurves()` function and save the result, which is a data frame. Below we use `subset()` to compare expected probability of failure at 10 weeks between the two rx groups. Probability of "survival" (staying in remission) for those in the treatment group is about 0.71, versus 0.46 in the placebo group.

```{r}
adjcurve <- surv_adjustedcurves(m2, variable = "rx", data = d)
subset(adjcurve, time == 10)
```

## CODE ALONG 4

(1) Fit a Cox PH model to the WHAS data using gender, age, hr (initial heart rate, bpm), and bmi.


(2) Assess the proportional hazards assumption of the model.


(3) What's the interpretation of the age coefficient?



## The Stratified Cox Model

Let's add the sex variable to our model. It is binary (1 = male, 0 = female). Perhaps our analysis plan specified that we would adjust for sex. This means retaining the sex variable regardless of significance. Below the coefficient is slightly positive, suggesting a higher hazard of failure for males. But the standard error is large and the z statistic is small. We're not sure what effect sex has on time to failure. Nevertheless we opt to keep it in the model.

```{r}
m3 <- coxph(Surv(survt, status) ~ rx + logwbc + sex, data = d)
summary(m3)
```

Now let's assess the proportional hazards assumption for all variables. The null is all variables satisfy the assumption. It appears we have evidence against the null for the sex variable.

```{r}
cox.zph(m3, global = FALSE)
```

Likewise the plot shows a curvy line. We index with `[3]` to see just the plot for sex. (sex is the third variable in the model.)

```{r}
plot(cox.zph(m3)[3])
```

One way to address a violation of the proportional hazards assumption is with a _stratified Cox model_. The general idea is to group the data according to the strata of a variable that violates the assumption, fit a Cox PH model to each strata, and combine the results into a single model. The coefficients of the remaining variables are _assumed to be constant across strata_. A drawback of this approach is the inability to examine the effects of the stratifying variable. Stratification is most natural when a covariate takes on only a few distinct values (like sex), and when the effect of the stratifying variable is not of direct interest.

We can implement a stratified Cox model by wrapping the variable to stratify on in the `strata()` function. Notice that sex is no longer reported in the model. Since we stratified on it, we do not estimate its effect. Also, the effects of rx and logwbc are assumed to be the same for males and females. This is called the _no-interaction assumption_.

```{r}
m4 <- coxph(Surv(survt, status) ~ rx + logwbc + strata(sex), data = d)
summary(m4)
```

We can test the no-interaction assumption by fitting a new model that allows the stratification to interact with the other variables, and then comparing this more complex model to the previous no-interaction model using a likelihood ratio test via `anova()`. The null is no difference in the models.

```{r}
m5 <- coxph(Surv(survt, status) ~ (logwbc + rx) * strata(sex), 
            data = d)
summary(m5)
```

Compare the interaction model with the no-interaction model.

```{r}
anova(m4, m5)
```

It appears the simpler no-interaction model is sufficient.

Finally, we can re-check the proportional hazards assumption of the stratified model.

```{r}
cox.zph(m4)
```


## CODE ALONG 5

(1) Fit a Cox PH model to the WHAS data using gender, age, hr (initial heart rate, bpm), and bmi, stratified on year (Cohort Year).


(2) Test the no-interaction assumption by fitting a new model that allows the stratification on year to interact with the other variables.


(3) Do we need to stratify on year?



## Regression diagnostics

When it comes to survival analysis, it's important to check if any subjects have high _leverage_ or _influence_. 

- Leverage: unusually large or small variables relative to rest of data
- Influence: removal would significantly change the model coefficients

High leverage itself is not necessarily a concern, however a subject with leverage may influence the estimation of model coefficients.

To assess leverage, we can use _score residuals_. Each subject in a Cox PH model will have a _separate score residual for each variable_ in the model. The larger a score residual for a particular variable (in absolute value), the larger the "leverage".

We can extract score residuals using the `residual()` function with `type = score`. Below we request score residuals for model m4. Notice there is a separate residual for each variable in the model.

```{r}
r <- residuals(m4, type = "score")
head(r)
```

For continuous variables, we can plot the score residual versus the variable. Below we check the logwbc residuals. We appear to have four subjects with unusually large residuals (in absolute value).

```{r}
plot(r[,"logwbc"] ~ d$logwbc)
```

For categorical variables, we can create boxplots of the score residuals versus the variable. Again we have a few outlying subjects we should investigate.

```{r}
boxplot(r[,"rxplacebo"] ~ d$rx)
```

For logwbc, we could check which subjects have a score residual less than -0.5

```{r}
i <- which(r[,"logwbc"] < -1) # select rows
i
```

Likewise we can check the rx score residuals.

```{r}
j <- which(r[,"rxplacebo"] < -0.4)
j
```

Subject 19 appears in both results. We can inspect the records for these subjects as follows. The `union()` function basically prevents subject 19 from appearing twice below.

```{r}
d[union(i,j),]
```

None of these subjects seem to have highly unusual values relative to the rest of the data. 

To assess influence, we can use _scaled score residuals_, also known as "dfbetas". Each subject in a Cox PH model will have a _separate dfbeta for each variable_ in the model. The larger a dfbeta for a particular variable (in absolute value), the larger the "influence".

We can extract score residuals using the `residual()` function with `type = dfbeta`. Below we request dfbetas for model m4. Again, notice there is a separate residual for each variable in the model. 

```{r}
r2 <- residuals(m4, type = "dfbeta")
head(r2)
```

These residuals are simply the score residuals multiplied by the model covariance matrix, which can be obtained with the function `vcov()`.

```{r}
# Use %*% for matrix multiplication
head(residuals(m4, type = "score") %*% vcov(m4)) 
```

Again we can use scatter plots and box plots to investigate influence.

```{r}
plot(r2[,2] ~ d$logwbc)
```

Four subjects seem to stand out:

```{r}
which(r2[,2] < -0.1)
```


```{r}
boxplot(r2[,2] ~ d$rx)
```

Three subjects appear to have outlying dfbeta values.

```{r}
which(r2[,1] < -0.1)
```

Once again subject 19 appears in both residual sets as having a large residual relative to the rest of the data.

To find out how variables are influencing a model, we need to drop them and refit the model. With such a small data set, we would want to think carefully before dropping subjects. However, let's say we want to see how much subject 19 is influencing the model. We can refit the model without subject 19 and calculate the change in the rx coefficient.

```{r}
m4a <- coxph(Surv(survt, status) ~ rx + logwbc + strata(sex), 
             data = d[-19,])
coef(m4a)["rxplacebo"]/coef(m4)["rxplacebo"]
```

With subject 19 removed from the model, the coefficient estimate for rx increases by almost 15%, which makes the treatment look even better. This doesn't mean we should drop the subject. It just helps us understand the impact a single subject has on our model. The decision to drop subjects should be carefully considered and not made based on just residual values.

## CODE ALONG 6

In the previous code along we fit the following model:

```{r}
whas_m3 <- coxph(Surv(time, fstat) ~ gender + age + hr + bmi + year, 
                 data = whas)
summary(whas_m3)
```

(1) Check the bmi score residuals for subjects with high leverage.



(2) Check the bmi dfbetas for subjects with high influence.



(3) Investigate any subjects that are turned up in these assessments and see how their removal from the model changes the results.



## The Extended Cox Model

We sometimes have variables that change over time (age, weight, status, etc). To accommodate these variables we can no longer assume proportional hazards. We must _extend_ the Cox model.

Time-dependent variables are usually coded as _intervals of time_. This is sometimes called the _counting process method_. For example:

```
    subject time1 time2 death  lab_value
1         1     0    90     0        0.7
2         1    90   160     0        1.9
3         1   160   220     1        1.3
```

Note: you cannot interpolate time-dependent values. For example, we cannot suppose the lab value is 1.3 at day 45.

The classic example of a Cox PH model with a time-dependent variable is the Stanford Heart Transplant study. 103 patients were enrolled in a program to receive a heart transplant between 1967 and 1974. Of interest is survival time while waiting for a heart transplant and after receiving a transplant. This data is available with the {survival} package as `jasa1`. The time-dependent variable is transplant. Notice subject 4 was observed for 35 days, at day 35 received a transplant, and died at day 38. (age = age - 48; year = years after 1 Nov 1967; surgery = prior bypass surgery (1 = yes))

```{r}
data(heart, package="survival")
head(jasa1, n = 10)
```

To include a time-dependent variable in a Cox PH model, we need to _include the time interval_ in the `Surv()` function. Notice below we have `Surv(start, stop, event)`. Here model the hazard of death as a function of whether the subject had a transplant, whether they had prior bypass surgery, and their age.

```{r}
tdm <- coxph(Surv(start, stop, event) ~ transplant + surgery + age,
             data= jasa1)
summary(tdm)
```

According to this simplistic model, prior bypass surgery seems to lower the hazard of death when while waiting for/after receiving a heart transplant by about 54%.



## Wrap-up

This was a short introduction to survival analysis. There is much we did not cover, such as:

- Additional wrinkles to modeling time-dependent covariates
- Right-censored survival times
- Parametric regression models
- Recurrent event models
- Frailty models (mixed-effect models for survival analysis)
- Competing risks models

Hopefully this workshop will give you the foundation to continue learning more about survival analysis.

## References

- Allison, P. (2010). Survival Analysis Using SAS, 2nd Ed. SAS Publishing.
- Harrell, F. (2015). Regression Modeling Strategies, 2nd Ed. Springer. 
- Hosmer, D., Lemeshow, S., and May, S. (2008). Applied Survival Analysis, 2nd Ed. Wiley.
- Kleinbaum, D. and Klein, M. (2005). Survival Analysis, a Self-Learning Text, 2nd Ed. Springer.




